---
title: "The Perfect Brew: Analyzing Coffee Ratings"
author: "Group C (Nabiha Tahsin, Louis Lin, Kinglee Tham)"
date: "4/1/25"
output: pdf
---

```{r, echo = FALSE, warning = FALSE, message = FALSE, include=FALSE}
library(ggformula)
library(mosaic)
library(readr)
library(gridExtra)
library(tidyverse)
library(GGally)
library(broom)
library(grid)
library(pracma)

coffee_data <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/06bca976827f922af8c1f62db67e6c1a7f0dcd4b/data/2020/2020-07-07/coffee_ratings.csv")
coffee_data <- coffee_data |>
  filter(total_cup_points > 1) |>
  select(total_cup_points, aftertaste, aroma, body, flavor, moisture, species)
```

# Project Aims and Data Source

This study aims to identify the most significant factors contributing to total cup points, a standardized coffee quality rating developed by the Coffee Quality Institute. Specifically, we investigate whether attributes such as flavor or aroma strongly influence overall ratings, and whether certain coffee species consistently receive higher evaluations than others. The analysis is based on a dataset of 1,340 digitized coffee reviews collected by James LeDoux in January 2018 through web scraping the publicly available Coffee Quality Database on Github. The “coffee_ratings.csv” data set contains ratings of different coffee beans from various locations and species. All beans were professionally rated on an overall 0 to 100 scale (total_cup_points) and on various factors such as aroma, sweetness, flavor, etc. While the database is continuously updated, this dataset represents a static snapshot from January 2018 and does not include data collected afterward. Our dataset includes coffee beans sourced globally, allowing us to draw generalizable conclusions about the factors that influence overall coffee quality. These insights can guide producers in optimizing desirable traits and support consumers in making informed, quality-based purchasing decisions.

# Variable Descriptions

Our response variable is *total cup points* (total_cup_points), which is an overall rating of the coffee beans, and it goes from a scale of 0 (worst) to 100 (best) points.

The six total explanatory variables used in this study include five quantitative variables (*aftertaste, aroma, body, flavor, moisture*) and one categorical variable (*species*):

*Aftertaste* (rated on a scale of 0 to 10, with 0 being the worst and 10 being the best) measures how long the coffee's flavor lingers after swallowing, affecting overall experience. 

*Aroma* (rated on a scale of 0 to 10, with 0 being the worst and 10 being the best) is the smell of the brewed coffee, which is crucial for consumer perception and enjoyment. 

*Body* (rated on a scale of 0 to 10, with 0 being the worst and 10 being the best) describes the texture and mouthfeel of coffee, including oiliness and weight. 

*Flavor* (rated on a scale of 0 to 10, with 0 being the worst and 10 being the best) is a key factor in overall quality, rating the intensity and pleasantness of taste. 

*Moisture* (measured as a percentage of the total weight of the sample) describes the moisture content of beans, which can impact freshness and roasting quality.

*Species* can be either Arabica or Robusta. Arabica is often associated with higher quality and more complex flavors, while Robusta tends to be stronger and more bitter.

# **Univariate Analyses**

One observation seemed to be an error since it was rated 0 in every quantitative variable, while the second-lowest-ranked bean scored 59.83 cup points and had ratings for aftertaste, aroma, body, and flavor between 6 to 8. We filtered the data set to remove the outlier.

```{r, echo = FALSE, warning = FALSE, message = FALSE}
# Create individual histograms
p1 <- gf_density(~ total_cup_points, data = coffee_data, binwidth = 0.3, fill = "lightblue") %>%
  gf_labs(title = "Distribution of\nTotal Cup Points", x = "Rating", y = "Count")

p3 <- gf_density(~ aftertaste, data = coffee_data, binwidth = 0.3, fill = "lightblue") %>%
  gf_labs(title = "Distribution of\nAftertaste ", x = "Rating", y = "Count")

p4 <- gf_density(~ aroma, data = coffee_data, binwidth = 0.3, fill = "lightblue") %>%
  gf_labs(title = "Distribution of\nAroma ", x = "Rating", y = "Count")

p6 <- gf_density(~ body, data = coffee_data, binwidth = 0.3, fill = "lightblue") %>%
  gf_labs(title = "Distribution of\nBody ", x = "Rating", y = "Count")

p7 <- gf_density(~ flavor, data = coffee_data, binwidth = 0.3, fill = "lightblue") %>%
  gf_labs(title = "Distribution of\nFlavor ", x = "Rating", y = "Count")

p8 <- gf_density(~ moisture, data = coffee_data, binwidth = 0.01, fill = "lightblue") %>%
  gf_labs(title = "Distribution of\nMoisture", x = "Moisture (%)", y = "Count") %>%
  gf_refine(scale_x_continuous(labels = percent_format(accuracy = 1)))

# Bar chart for counts of species
p9 <- gf_bar(~ species, data = coffee_data, fill = "lightblue") |>
   gf_labs(title = "Bar Chart of Species",
       	x = "Species", y="Count")
```

```{r, echo = FALSE, warning = FALSE, message = FALSE, fig.width = 12, fig.height = 4}
# Plot the histograms and bar chart
grid.arrange(
  p9,          
  grid.arrange(p1, p3, p4, p6, p7, p8, nrow = 2),
  widths = c(1, 3)  
)
# Tally by species
  tally(coffee_data$species)
```

```{r, echo = FALSE, eval = FALSE}
# Don't evaluate this in the render
  summary(coffee_data)
  sapply(coffee_data, function(x) if(is.numeric(x)) sd(x, na.rm = TRUE))
```

### Bar Chart

The bar chart of species indicates that there are 1310 observations of Arabica beans and only 28 observations of Robusta beans. Given that we have so few observations of Robusta included in our data set, it would not make a balanced analysis to include Robusta beans in our data set; we don't have enough data to predict values for Robusta. **As a result, we filtered our data set to include only Arabica beans and removed the *species* categorical variable.**

```{r, echo = FALSE}
coffee_data <- coffee_data |>
  filter(species == "Arabica")
```

### Histograms

The density plot for *total cup points* (mean = 82.15, SD = 2.68), *aftertaste* (mean = 7.407, SD = 0.35), *aroma* (mean = 7.572, SD = 0.32), body (mean = 7.523, SD = 0.31), and flavor (mean = 7.526, SD = 0.34) are all normally distributed with no obvious outliers or skewness. The density for *moisture* looks bimodal with one smaller peak around 0% and one peak around 10-12%.

# Bivariate Analyses

### Quantitative vs. Categorical Variables

Since we removed *species* as a variable, we did not create boxplots of *total_cup_points* vs. quantitative predictors, with species as a factor.

### Quantitative vs. Quantitative Variables

To check for bivariate relationships between quantitative variables, we created a scatterplot of every combination of quantitative variables and derived their respective correlation coefficients.

Linearity Condition: The scatterplots of *aftertaste, aroma, body,* and *flavor* on *total_cup_points* appear linear, so they fit the linearity condition. The aforementioned quantitative predictors have moderately positive linear relationships with *total_cup_points*, with their correlation coefficients respectively being 0.812, 0.686, 0.628, and 0.827. *Moisture* does not appear to have a linear relationship with *total_cup_points* or any other predictor variables, indicating that it does not pass the linearity condition. There are also two clusters in all *moisture* scatterplots plots, which reflected the bimodal distribution of *moisture*.

Equal Variance Condition: We plotted the residuals between our quantitative predictors (respectively, *aftertaste, aroma, body, flavor, moisture respectively*) and *total_cup_points*. The relationship between all of our quantitative variables and total_cup_points looks heteroskedastic, as the spread of *total_cup_points* increases when *aftertaste, aroma, body,* and *flavor* are rated low. The relationship between moisture and total_cup_points has a butterfly shape, as the spread increases and decreases throughout different moisture ratings. None of our quantitative predictors fit the equal variance condition, so we transformed them.

```{r, echo = FALSE, include = FALSE}
# making tests mod so we can see equal variance of each quant pred vs total cup pts
aftertastemod <- lm(total_cup_points ~ aftertaste, data = coffee_data)
aromamod <- lm(total_cup_points ~ aroma, data = coffee_data)
bodymod <- lm(total_cup_points ~ body, data = coffee_data)
flavormod <- lm(total_cup_points ~ flavor, data = coffee_data)
moisturemod <- lm(total_cup_points ~ moisture, data = coffee_data)

# Make a resids plot for each individual quant predictor vs cup points
# Residual plots for aftertaste, aroma, body,flavor, moisture respectively
aftertasteresids <- mplot(aftertastemod, which = 1)
aromaresids <- mplot(aromamod, which = 1)
bodyresids <- mplot(bodymod, which = 1)
flavorresids <- mplot(flavormod, which = 1)
moistureresids <- mplot(moisturemod, which = 1)


```

```{r, echo = FALSE, include = FALSE}
# Transform variables
coffee_data <- coffee_data |>
  mutate(
    aftertaste_t = log10(aftertaste),
    aroma_t      = log10(aroma),
    body_t       = log10(body),
    flavor_t     = log10(flavor),
    moisture_t   = log10(moisture)
  )

# making tests mod so we can see equal variance of each quant pred vs total cup pts
aftertastemod_t <- lm(total_cup_points ~ aftertaste_t, data = coffee_data)
aromamod_t <- lm(total_cup_points ~ aroma_t, data = coffee_data)
bodymod_t <- lm(total_cup_points ~ body_t, data = coffee_data)
flavormod_t <- lm(total_cup_points ~ flavor_t, data = coffee_data)


# Make a resids plot for each individual quant predictor vs cup points
# Residual plots for aftertaste, aroma, body,flavor, moisture respectively
aftertasteresids_t <- mplot(aftertastemod_t, which = 1)
aromaresids_t <- mplot(aromamod_t, which = 1)
bodyresids_t <- mplot(bodymod_t, which = 1)
flavorresids_t <- mplot(flavormod_t, which = 1)

```

```{r, echo = FALSE, fig.height = 6, fig.width = 15}
# Organize all resid plots
# before transformation
grid.arrange(aftertasteresids, aromaresids, bodyresids, flavorresids, moistureresids, aftertasteresids_t, aromaresids_t, bodyresids_t, flavorresids_t, nrow = 2)
```

Independence Condition: Each bean was rated independently regardless of other beans, so it meets the independence condition.

No Outliers Condition: There are some outliers in the scatterplots between each quantitative predictor and total_cup_points, so it does not meet the no outliers condition. In general, there are a lot of outliers when the fitted value is low.

Collinearity Condition: Given that aftertaste, aroma, body, and flavor have moderate correlations between each other, we would like to check the variation inflation factor (VIF) for possible concerns with multicollinearity. However, we will need to construct a model to check VIF, so we will return to this after building a model.

No transformation could make *moisture'*s distribution unimodal. Also, it met the fewest of the conditions and had the weakest correlation with *total_cup_points*, so **we removed moisture from the possible variables**.

```{r}
quant_vars <- coffee_data |>
  select(aftertaste, aroma, body, flavor, moisture, total_cup_points)

# Create a correlation matrix
cor_matrix <- cor(quant_vars, use = "complete.obs")
print(cor_matrix)

# Another matrix
ggpairs(
  quant_vars,
  lower = list(continuous = wrap("points", color = "lightblue", size = 0.8)),
  diag = list(continuous = wrap("densityDiag", fill = "lightblue")),
  upper = list(continuous = wrap("cor", size = 4))
)
```

# Step 3: Building Models

We ran an automated variable selection approach to select the best subset of predictor variables to include in models of different sizes. Based on the best subsets output, we selected the four-predictor model with aftertaste, body, flavor, and aroma as predictors after including only significant variables and balancing high parsimony, high adjusted R2 (0.737), and low Mallows’ Cp (7.00).

```{r}
library(leaps)
best_subset <- regsubsets(total_cup_points ~ aftertaste + aroma + body + flavor + moisture,  data = coffee_data, nbest = 1)
with(summary(best_subset), data.frame(rsq, adjr2, cp, rss, outmat))
```

After analyzing the correlation matrix and VIF values, we will remove explanatory variables that are collinear with each other (with VIFs higher than 5) and prioritize explanatory variables with the highest correlation to total_cup_points.

```{r}
subset_model <- lm(total_cup_points ~ aftertaste + aroma +  body + flavor, data = coffee_data)

library(car)
vif(subset_model)
```

Though aftertaste and flavor's VIF values appear to be close to 5, they do not reach or go beyond 5, so there is no significant issue of multicollinearity in our selected variables.

After building the model, we will check conditions of equal variance and normality of errors using residual plots and QQ plots, respectively.

```{r}
subset_model <- lm(total_cup_points ~ aftertaste + aroma +  body + flavor, data = coffee_data)

subset_mod_resid <- mplot(subset_model, which = 1)
subset_mod_normality <- mplot(subset_model, which = 2)


# Organize all resid plots
grid.arrange(subset_mod_resid, subset_mod_normality, nrow = 2)
```

The equal variance condition appear to be met. However, the lower tail of the Q-Q plot deviates away from the straight line. We will try removing outliers and looking at the conditions again to see if it improves the conditions. To determine outliers that may be influential, we will use Cook's distance.

```{r}
cooksD <- cooks.distance(subset_model)

plot(cooksD, type = "h",
     main = "Cook's Distance",
     ylab = "Cook's D", xlab = "Observation Index",
     col = "lightblue", lwd = 2)


```

```{r}
# Three-most unusual points in the data set
cooksD[c(1310, 1309, 1296)]
```

It appears that none of the points have a Cook's distance greater than 0.5 (for moderate influence) or 1.0 (for high influence), so we should not remove any coffee bean observation.

We also take a look at added variable plots to check if each of the predictors influence our response, total_cup_points, taking into account the effect of other variables

```{r}
car::avPlots(subset_model) 
```

The added-variable plots show that points 1309 and 1310 are outliers, but we've already calculated Cook's distances for these points and decided not to remove them based on the cutoff. The predictor variables seem to have positive relationship with the total_cup_points, though the slopes are not very steep (but also not flat!).

Lastly, to compare multiple models, we bring in the kitchen sink model to relate with the subsets model.

```{r}
kitchen <- lm(total_cup_points ~ aftertaste+aroma+flavor+body+moisture, data= coffee_data)
anova(subset_model, kitchen)
```

```{r}
msummary(kitchen)
```

It seems that moisture does not add enough information to be a significant predictor of total cup points (p = 0.08 \> 0.05), so our subsets model is the best model we can use.

Finally, with our list of explanatory variables that pass the conditions and significant interactions, we created our final multiple linear regression.

```{r}

final_model <- lm(total_cup_points ~ flavor + aftertaste + body + aroma, data = coffee_data)
summary(final_model)

```

```{r}
# ggplot(coffee_data, aes(x = flavor, y = total_cup_points, color = species)) +
#   geom_point(alpha = 0.7) +
#   geom_smooth(method = "lm", se = FALSE) +
#   labs(title = "Flavor vs Total Cup Points by Species",
#        x = "Flavor",
#        y = "Total Cup Points") +
#   theme_minimal()
```
